### 2025-08-18 chapters/introduction.tex
5.3 complete

\chapter{Introduction}
\label{chap:intro}
Modern computing increasingly relies on platforms that operate in potentially untrusted environments. These include public cloud infrastructures, multi-tenant servers, and distributed edge devices, where sensitive data and critical computations may be exposed to compromised software stacks or malicious actors. In such contexts, ensuring the confidentiality and integrity of both data and execution is a fundamental requirement. Conventional security mechanisms, such as access control and encryption, are often insufficient in scenarios where the underlying operating system, hypervisor, or even firmware may be compromised.

To address these challenges, Trusted Execution Environments (TEEs) have emerged as a compelling solution. TEEs provide hardware-assisted isolation mechanisms that enable the execution of code in a protected context, referred to as an enclave. This enclave is designed to be resilient to attacks originating from untrusted system components, including the operating system and other user-space applications. By establishing a secure boundary around sensitive code and data, TEEs allow developers to implement trustworthy services even on compromised hosts.

Several commercial implementations of TEEs exist today, with Intel Software Guard Extensions (SGX) and ARM TrustZone being the most prominent examples. These technologies have demonstrated the practical viability of hardware-enforced isolation, and they have been integrated into a variety of real-world applications, from digital rights management to secure machine learning inference. However, despite their utility, these solutions are inherently proprietary and closed-source. This lack of transparency imposes significant constraints on researchers, developers, and system architects who wish to investigate alternative TEE designs, customize security mechanisms, or perform formal verification. Moreover, the architectural rigidity of these platforms limits their adaptability to emerging use cases, especially in academic or experimental contexts.

To overcome these limitations, the Keystone framework introduces an open-source, modular TEE architecture built atop the RISC-V instruction set architecture (ISA). RISC-V itself is an open, extensible ISA that has gained significant traction in both academic and industrial settings. Keystone leverages RISC-V's flexibility to enable fine-grained control over enclave behavior and to support a minimal, verifiable trusted computing base (TCB). The framework allows researchers and developers to explore custom enclave policies, experiment with hardware-software co-design, and perform reproducible evaluations in a controlled setting.

This thesis presents a detailed investigation into the Keystone TEE framework. It explores Keystone's architectural foundations, its use of RISC-V hardware features such as Physical Memory Protection (PMP), and its support for dynamic enclave management. Special attention is paid to the performance implications of enclave isolation, as measured through controlled benchmarking in a virtualized environment. By evaluating Keystone under varying execution conditions, this work seeks to provide a rigorous assessment of its efficiency and suitability for secure computation on emerging RISC-V platforms.

@@

This thesis makes significant contributions to the study and evaluation of Trusted Execution Environments (TEEs), with a particular focus on the Keystone framework. The research encompasses architectural analysis, emulation-based deployment, and rigorous empirical performance evaluation.

Firstly, a comprehensive architectural analysis of the Keystone TEE framework is presented. This includes an in-depth discussion of its core components, such as the Security Monitor (SM), enclave runtime environment, and user-level application design. The role of RISC-V’s privilege hierarchy and Physical Memory Protection (PMP) in enforcing strict isolation guarantees is thoroughly elaborated, providing a holistic understanding of Keystone’s security architecture.

Secondly, the thesis conducts a rigorous comparative evaluation of Keystone alongside widely deployed commercial TEEs, specifically Intel Software Guard Extensions (SGX) and ARM TrustZone. This analysis highlights fundamental architectural distinctions, including enclave creation models, privilege separation strategies, scalability, Trusted Computing Base (TCB) size, and extensibility. Such comparison contextualizes Keystone’s design philosophy and underscores its advantages in openness and modularity.

Furthermore, the practical deployment of Keystone within a fully virtualized environment is demonstrated using a RISC-V-specific fork of the QEMU emulator. This approach facilitates development and testing independent of physical RISC-V hardware. The thesis also documents the toolchain configuration, build process, and debugging workflow, establishing a reproducible methodology for future research.

To quantify the runtime overhead associated with enclave-based execution, Keystone enclaves are evaluated using two established synthetic benchmarks: Dhrystone and CoreMark. Key performance metrics—including execution time, Dhrystones per second (DMIPS), iterations per second (IPS), and variability across multiple runs—are measured to assess the cost of enclave isolation and security enforcement.

Finally, the computational behavior of benchmark applications within the enclave context is analyzed to provide empirical insights into how TEE-induced isolation affects instruction throughput and memory access patterns. These findings illuminate the inherent trade-offs between performance and security in enclave-based execution models.

@@

The field of Trusted Execution Environments has been shaped by a combination of commercial deployments and academic proposals, each addressing different aspects of secure computing. This section situates the present work within the broader landscape of TEE research and implementation.

Intel SGX is one of the most well-known commercial TEE solutions. It introduces user-space enclaves with hardware-enforced isolation and memory encryption. SGX supports fine-grained protection mechanisms and remote attestation, making it suitable for confidential computing scenarios. However, SGX is hindered by several limitations, including rigid enclave memory constraints, lack of supervisor-mode support, and a closed-source implementation that precludes low-level architectural exploration or modification. These issues restrict its applicability in research and experimental systems development.

ARM TrustZone represents a different paradigm, based on the separation of execution into a Secure World and a Normal World. While TrustZone enables trusted operating systems and supports a variety of secure services, it suffers from scalability issues, particularly in multi-core environments where interrupt handling becomes a bottleneck. Moreover, TrustZone's coarse-grained isolation and reliance on a monolithic Secure World kernel contribute to a large TCB, which poses challenges for security assurance and formal verification.

In the academic domain, projects such as Sanctum and Sancus have contributed lightweight and formally verifiable secure execution frameworks. Sanctum builds on RISC-V to enforce secure memory isolation and secure multiplexing of shared hardware resources. Sancus targets embedded systems and emphasizes minimalism and provable security properties. While these efforts have made valuable theoretical contributions, they are often constrained by limited platform support and lack the general-purpose capabilities required for broader deployment or benchmarking.

Keystone addresses the limitations of both commercial and academic TEEs by adopting a fully open-source design built on RISC-V. It supports dynamic enclave creation, configurable runtime environments, and an extensible architecture suitable for incorporating advanced security primitives. Keystone’s modularity, support for supervisor-level enclave components, and compatibility with standard Linux-based host systems make it especially well-suited for research, prototyping, and educational use. This thesis builds upon this foundation by exploring the architectural, practical, and performance aspects of deploying Keystone in a virtualized environment, thereby contributing new empirical insights and documentation to the growing body of work on open TEEs.

@@

This thesis is structured to progressively build an understanding of Trusted Execution Environments (TEEs) with a focus on the Keystone framework, moving from conceptual foundations to empirical evaluation and practical system design insights.

This thesis is organized as follows: Chapter~\ref{chap:intro} introduces the motivation for secure computation in untrusted environments, outlines the core contributions of this work, and surveys relevant literature in the domain of Trusted Execution Environments (TEEs). Chapter~\ref{chap:background} presents the technical foundations and system architecture of the Keystone framework, including a comparative analysis with Intel SGX and ARM TrustZone that highlights key design trade-offs and performance considerations. Chapter~\ref{chap:methodology} describes the experimental setup, detailing the virtualized testbed, benchmark selection, parameter tuning, and data collection methods employed to ensure rigorous and reproducible evaluation. Chapter~\ref{chap:results} reports empirical findings and analyzes the impact of system parameters such as CPU core count, memory allocation, and cache size on enclave performance, along with associated overheads and insights. Chapter~\ref{chap:recommendations} synthesizes these results into practical configuration guidelines for deploying Keystone TEEs across diverse workload scenarios and identifies architectural bottlenecks for future improvements. Finally, Chapter~\ref{chap:conclusion} summarizes the main contributions and discusses their implications for secure system design, outlining promising directions for further research including hardware extensions and formal verification.

>>>>

\chapter{Introduction}
\label{chap:intro}

In the rapidly evolving landscape of modern computing, an undeniable shift has taken place over the past decade. Computing no longer happens exclusively within the safe confines of dedicated, physically secured machines operated by trusted personnel. Instead, it is increasingly conducted across a wide array of environments that range from large-scale, multi-tenant cloud platforms to geographically dispersed edge devices deployed in untrusted, sometimes hostile, settings. These emerging scenarios offer unparalleled scalability, accessibility, and flexibility but simultaneously pose profound new challenges to the foundational assumptions upon which traditional system security models have long been based.

Consider, for example, a cloud computing environment where organizations lease computational resources from a third-party provider. While this model delivers significant economic and operational advantages, it necessarily requires placing sensitive data and critical computations on infrastructure controlled, at least in part, by entities outside the organization’s direct control. In such environments, the software stack—from firmware and bootloaders to the operating system and hypervisor—may be susceptible to compromise through vulnerabilities, misconfigurations, or even insider threats. Similarly, edge devices deployed in remote locations may operate without continuous physical supervision, making them vulnerable to physical tampering or software-based attacks. In all these cases, the question arises: how can one guarantee that sensitive computations execute correctly and confidentially, even when the underlying platform may be compromised or untrusted?

Traditional security approaches, such as enforcing access controls, employing encryption for data at rest or in transit, or leveraging virtualization-based isolation, offer important protections but are often insufficient in these adversarial contexts. These mechanisms fundamentally rely on the integrity of the underlying system software to enforce security policies. Once that trust is broken—for example, if a privileged kernel or hypervisor is subverted—malicious actors may gain unauthorized access to secrets, manipulate computations, or subvert the system’s intended behavior. As a result, there has been a growing recognition of the need for a fundamentally different approach: one that can provide strong guarantees about the confidentiality and integrity of code and data, even in the face of potentially compromised system components.

This pressing need has catalyzed the development of Trusted Execution Environments (TEEs), a class of hardware-assisted security mechanisms designed to establish a secure and isolated environment within an untrusted platform. A TEE creates a trusted “enclave” or protected region of execution where sensitive code and data reside, shielded from the rest of the system, including the operating system, hypervisor, and other software running on the host. By leveraging hardware features that enforce strict isolation and memory protection boundaries, TEEs aim to provide a sanctuary of trust—a secure vault in which computations can proceed without fear of interference or observation by unauthorized parties.

The promise of TEEs is compelling: they enable the secure execution of sensitive workloads such as cryptographic key management, digital rights enforcement, confidential cloud computing, and privacy-preserving data analytics. Over the past several years, commercial implementations have emerged, most notably Intel’s Software Guard Extensions (SGX) and ARM’s TrustZone technology, each offering unique approaches to enclave creation and hardware isolation. Intel SGX, for example, introduces user-space enclaves with hardware-enforced memory encryption and attestation capabilities, allowing remote verification of enclave authenticity. ARM TrustZone takes a different approach, partitioning the processor into Secure and Normal worlds to separate trusted code from the rest of the system.

Despite their successes and practical deployments, these commercial TEEs come with a number of significant limitations that hinder their applicability in research, experimental settings, and use cases requiring flexibility and transparency. Both SGX and TrustZone are proprietary technologies with closed-source implementations, limiting independent analysis, verification, and modification. Their architectural designs are relatively rigid, optimized for particular use cases, and do not easily accommodate novel security policies, alternative enclave management strategies, or supervisor-mode enclave execution. Moreover, their Trusted Computing Bases (TCBs) can be substantial, complicating efforts at formal verification and security assurance. These challenges collectively motivate the exploration of open, extensible TEE architectures that can empower researchers and practitioners to experiment, customize, and extend trusted execution mechanisms without the constraints imposed by proprietary solutions.

It is within this context that the Keystone framework has emerged as a pioneering open-source TEE architecture built atop the RISC-V instruction set architecture (ISA). RISC-V, an open and extensible ISA, has attracted substantial attention in academia and industry alike for its flexibility, simplicity, and openness. Keystone leverages these advantages to provide a modular, fully open TEE framework that supports fine-grained control over enclave behavior and a minimal trusted computing base. Notably, Keystone harnesses RISC-V’s hardware capabilities such as Physical Memory Protection (PMP) to implement hardware-enforced isolation, enabling dynamic enclave management and runtime configurability rarely available in existing TEEs.

Keystone’s open design philosophy not only facilitates deep architectural exploration and formal verification but also enables the development of novel security primitives and policies tailored to emerging application domains. Its modularity supports supervisor-level enclave components, dynamic loading, and interaction with Linux-based host environments, making it well-suited for research, prototyping, and education. However, the flexibility offered by Keystone comes with inherent challenges. For instance, RISC-V’s PMP architecture, while powerful, imposes a limited number of protection entries, constraining the number of concurrent enclaves and complicating memory management. Additionally, the performance overhead induced by enclave isolation—especially when executing memory-intensive or computationally demanding workloads such as post-quantum cryptographic algorithms—necessitates thorough empirical evaluation to guide practical deployment and future hardware design.

This thesis undertakes a comprehensive investigation of the Keystone framework, focusing on its architectural foundations, practical deployment in a fully virtualized RISC-V environment, and performance characteristics under representative workloads. Through methodical benchmarking using standard synthetic tests and cryptographic workloads, this work illuminates the complex trade-offs between security, performance, and scalability inherent to Keystone’s design. The study further documents the intricacies of configuring and debugging Keystone enclaves in an emulator, addressing the challenges posed by hardware constraints such as PMP entry exhaustion. By providing a rigorous empirical characterization of Keystone’s runtime behavior and scalability limits, this thesis contributes valuable insights to the open TEE community and lays the groundwork for future enhancements in RISC-V based secure execution environments.

In the following sections, this introduction will first articulate the specific contributions of this thesis in detail. It will then situate this work within the broader context of related research and commercial TEE developments. Finally, an overview of the thesis structure will guide the reader through the chapters that follow, setting the stage for an in-depth exploration of Keystone and its capabilities.

@@

This thesis makes several key contributions that advance the understanding and evaluation of Trusted Execution Environments, particularly focusing on the open-source Keystone framework and its RISC-V hardware foundation.

\begin{itemize}
    \item \textbf{Comprehensive Architectural Analysis:} A detailed exposition of Keystone’s design is presented, highlighting its Security Monitor (SM), enclave runtime, and user-level application interactions. This includes an exploration of how RISC-V’s privilege levels and Physical Memory Protection (PMP) mechanisms are employed to establish and enforce robust isolation boundaries between enclaves and untrusted system components.

    \item \textbf{Comparative Evaluation with Commercial TEEs:} The thesis performs a thorough comparative study positioning Keystone alongside Intel SGX and ARM TrustZone. This analysis elucidates key architectural distinctions, including enclave lifecycle management, privilege separation, Trusted Computing Base size, scalability, and extensibility. This contextualization highlights Keystone’s unique advantages in openness, modularity, and adaptability for research and experimentation.

    \item \textbf{Virtualized Deployment and Toolchain Documentation:} A fully virtualized deployment of Keystone is demonstrated using a RISC-V-specific QEMU emulator. This approach allows experimentation independent of physical hardware availability. The thesis provides detailed documentation of the build environment, toolchain configuration, emulator setup, and debugging workflows, establishing a reproducible methodology for future work.

    \item \textbf{Empirical Performance Evaluation:} Using established benchmarks such as Dhrystone and CoreMark, alongside a computationally demanding post-quantum cryptographic workload (Kyber), the thesis quantifies runtime overheads incurred by enclave isolation. It explores how varying system parameters—such as CPU core count, memory allocation, and cache sizes—affect enclave performance and stability.

    \item \textbf{Analysis of Hardware-Induced Scalability Constraints:} The limited number of PMP entries and its impact on enclave concurrency and lifecycle management is carefully analyzed. The thesis identifies bottlenecks and proposes strategies to mitigate the effects of PMP exhaustion on system scalability and enclave deployment.

    \item \textbf{Identification and Documentation of Runtime Failure Modes:} Through extensive testing, the thesis uncovers failure scenarios arising from PMP limitations and enclave memory protection policies, contributing practical insights and guidance to the Keystone community for improving robustness and fault tolerance.
\end{itemize}

@@

Trusted Execution Environments (TEEs) have become a cornerstone for securing computation on untrusted platforms by isolating sensitive code and data from the rest of the system. The landscape of TEE technologies includes Intel SGX, ARM TrustZone, AMD SEV, and emerging open-source frameworks like Keystone, each offering distinct security guarantees, performance trade-offs, and hardware dependencies.

Intel SGX~\cite{akkram2020scientific, kumar2022sgxgauge} provides strong confidentiality and integrity guarantees via isolated enclaves backed by hardware-enforced memory encryption. However, SGX suffers from limited enclave page cache (EPC) size, causing significant performance degradation when working sets exceed EPC limits, as observed by Kumar et al.~\cite{kumar2022sgxgauge}. Moreover, the overheads in enclave transitions and paging have been documented in multiple studies~\cite{akkram2020scientific, krahn2020teemon}.

ARM TrustZone~\cite{suzaki2021ts, turn0search8} takes a different approach by partitioning the processor into secure and normal worlds, enabling lightweight isolation for Trusted Applications but with generally weaker isolation guarantees than SGX. Its performance impact tends to be lower, yet TrustZone’s capabilities and APIs are less standardized, which complicates broad adoption and benchmarking~\cite{turn0search9}.

AMD SEV and newer technologies such as SEV-SNP offer memory encryption and integrity protections for virtual machines, providing scalable enclave-like isolation for cloud workloads. Coppolino et al.~\cite{turn0search2_coppolino} experimentally evaluated SEV and TDX against SGX, revealing performance and transparency trade-offs that influence their suitability for different application scenarios.

In parallel, open-source efforts like Keystone~\cite{Lee2019, dayeol2019keystone} aim to democratize TEE development by providing a modular framework on RISC-V processors, supporting flexible enclave configurations and custom security policies. Research into Keystone’s memory models and sharing mechanisms~\cite{yu2022elasticlave, lee2022cerberus} has demonstrated improved performance and formally verified security properties, addressing common TEE bottlenecks such as enclave memory management overhead.

Benchmarking and performance measurement frameworks are crucial to evaluate these TEEs fairly and holistically. TS-perf~\cite{suzaki2021ts} offers a unified performance suite targeting multiple TEE architectures, including SGX, TrustZone, and Keystone, facilitating comparative studies. Other benchmarking efforts include SGXGauge~\cite{kumar2022sgxgauge}, which specifically assesses Intel SGX overheads under different workload characteristics, and traditional CPU benchmarks like Dhrystone~\cite{weiss2002dhrystone, york2002benchmarking} and CoreMark~\cite{gal2012exploring}, adapted for enclave contexts.

Continuous monitoring tools like TEEMon~\cite{krahn2020teemon} help identify runtime bottlenecks by tracking enclave performance metrics with modest overheads, enabling dynamic optimization and debugging.

Recent surveys~\cite{Survey2023, turn0search5} provide comprehensive overviews of TEE design choices, common pitfalls, and application domains, highlighting the need for standardized performance evaluations and security validations.

While the state of the art has advanced significantly, challenges remain in balancing security, performance, and programmability across diverse hardware platforms and threat models. This work builds on these prior efforts by focusing on comprehensive performance evaluation across multiple TEEs using realistic workloads, emphasizing the modularity and transparency advantages offered by Keystone and comparing them with established commercial solutions.

@@

This thesis is structured to progressively build the reader’s understanding of Trusted Execution Environments, with an emphasis on the Keystone framework and its RISC-V underpinnings. Each chapter deepens the narrative from foundational concepts to practical experimentation and critical analysis.

\begin{itemize}
    \item \textbf{Chapter~\ref{chap:intro}} lays the groundwork by introducing the motivation for secure computing in untrusted environments, detailing the contributions of this work, surveying relevant literature, and providing an overview of the thesis structure.

    \item \textbf{Chapter~\ref{chap:background}} presents essential background material on Trusted Execution Environments, the Keystone framework, and RISC-V architectural features such as Physical Memory Protection. It also introduces the Kyber post-quantum key encapsulation mechanism used as a cryptographic benchmark.

    \item \textbf{Chapter~\ref{chap:methodology}} describes the experimental setup, including the virtualized RISC-V testbed, benchmark selection criteria, parameter tuning, and measurement methodologies ensuring rigor and reproducibility.

    \item \textbf{Chapter~\ref{chap:benchmarking}} reports detailed empirical results exploring the performance impact of enclave isolation under different system configurations and workloads, identifying key overheads and architectural bottlenecks.

    \item \textbf{Chapter~\ref{chap:discussion}} synthesizes findings from prior chapters to discuss the trade-offs inherent in enclave design and execution, highlighting implications for future hardware and software secure computing architectures.

    \item \textbf{Chapter~\ref{chap:conclusion}} concludes the thesis by summarizing main contributions, reflecting on their broader significance, and outlining promising avenues for further research, including hardware extensions and formal verification approaches.
\end{itemize}

### 2025-08-18 chapters/system-architecture.tex
introduced latex diagrams
5.3 complete

@@

@@

\begin{itemize}
    \item \texttt{QEMU\_PORT} (default: 9821) — SSH port forwarding to access the QEMU guest.  
    \item \texttt{QEMU\_DBG\_PORT} (default: \texttt{QEMU\_PORT} + 1) — TCP port for GDB debugging.  
    \item \texttt{QEMU\_RAM\_SIZE} (default: 128M) — amount of emulated RAM.  
    \item \texttt{QEMU\_CORE\_COUNT} (default: 2) — number of virtual CPU cores.
\end{itemize}

@@

\begin{itemize}
    \item Instruction cache: \texttt{icachesize}, \texttt{iblksize}, \texttt{iassoc} — size, block size, and associativity.  
    \item Data cache: \texttt{dcachesize}, \texttt{dblksize}, \texttt{dassoc} — analogous data cache parameters.  
    \item Eviction policy: \texttt{evict=lru|rand|fifo}.  
    \item Logging limits: \texttt{limit=TOP\_N} for number of top thrashing instructions to record.  
    \item Number of cores monitored: \texttt{cores=N\_CORES}.
\end{itemize}

@@

\begin{itemize}
    \item \texttt{indcpa\_keypair} — Key pair generation for the IND-CPA variant.
    \item \texttt{indcpa\_enc} — Encryption (encapsulation) under IND-CPA assumptions.
    \item \texttt{indcpa\_dec} — Decryption (decapsulation) under IND-CPA assumptions.
    \item \texttt{kyber\_keypair} — Enhanced key pair generation supporting IND-CCA.
    \item \texttt{kyber\_encaps} — IND-CCA-compliant encapsulation.
    \item \texttt{kyber\_decaps} — IND-CCA-compliant decapsulation.
\end{itemize}

Each of these operations was executed across the five representative system configurations defined in Section~\ref{sec:param-variation}, which span a broad range of memory capacities, core counts, and execution modes. To ensure measurement accuracy and reduce variability from transient system states, ten iterations were run per configuration with the first iteration discarded as a warm-up phase. Execution timing was obtained using the enclave’s internal high-resolution timer API, capturing secure-world runtime exclusively. In parallel, external system-level metrics — such as CPU usage, memory footprint, and context switching counts — were gathered from the host environment to provide contextual insights into system overhead and resource utilization.

@@

\begin{itemize}
    \item Small caches (4--8\,KB) with low associativity and an LRU replacement policy
    \item Mid‑sized caches (8--16\,KB) with mixed associativity and FIFO or random replacement
    \item Large caches (32\,KB) with high associativity and FIFO eviction
\end{itemize}

@@

The primary performance metric collected was **total wall-clock execution time**, as measured from within the guest Linux environment using high-resolution timers (e.g., \texttt{clock\_gettime()}) or the Keystone timer API in enclave mode. For Dhrystone and CoreMark, benchmark-specific throughput metrics were also recorded, including Dhrystone MIPS (DMIPS) and iterations per second. These allowed for workload-specific efficiency comparisons across configurations.

In addition to these high-level metrics, detailed system-level runtime statistics were gathered using standard Linux facilities such as the \texttt{/proc} filesystem and the \texttt{getrusage()} system call. These included:

\begin{itemize}
    \item \textbf{User and system CPU time:} Total time spent in user-space and kernel-space respectively.
    \item \textbf{Maximum resident set size (RSS):} Peak memory consumption during execution.
    \item \textbf{Page fault statistics:} Number of minor (soft) and major (hard) page faults incurred during the benchmark.
    \item \textbf{Context switching behavior:} Number of voluntary and involuntary context switches, indicating scheduling and synchronization overhead.
\end{itemize}

>>>>

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=1.2cm and 1.5cm,
    every node/.style={
        draw,
        rounded corners,
        align=center,
        font=\small,
        minimum width=2.8cm,
        minimum height=1.2cm,
        fill=gray!15,
        drop shadow={shadow xshift=0.5ex,shadow yshift=-0.5ex,opacity=0.1}
    },
    keystone/.style={
        fill=white,
        thick,
        minimum width=3.5cm,
        drop shadow={shadow xshift=0.8ex,shadow yshift=-0.8ex,opacity=0.2}
    },
    connector/.style={
        ->,
        thick,
        >=stealth,
        line width=0.6pt
    }
]

% Nodes
\node (host) [fill=gray!30, minimum height=1.4cm] {Ubuntu 22.04 LTS\\ Host OS};
\node (qemu) [below=of host] {QEMU RV64GC Emulator};
\node (keystone) [keystone, below=of qemu] {Keystone Stack\\ \footnotesize Security Monitor, Runtime, Enclaves};
\node (virtio) [fill=gray!20, left=2cm of keystone] {VirtIO Devices\\ (net, storage)};
\node (plugin) [fill=gray!20, right=2cm of keystone] {Optional Cache Plugin};

% Edges
\draw[connector] (host) -- (qemu);
\draw[connector] (qemu) -- (keystone);
\draw[connector] (virtio) -- (keystone);
\draw[connector] (plugin) -- (keystone);

% Annotation
\node[font=\scriptsize, anchor=south west] at ([xshift=0.3cm,yshift=0.3cm]host.north west) {SSH/GDB access from Host};

\end{tikzpicture}
\caption{Overall virtualized testbed architecture illustrating the host OS, QEMU emulator, Keystone stack, and peripheral components such as VirtIO devices and optional cache plugin.}
\label{fig:virtualized_testbed_architecture}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=1.5cm,
    every node/.style={
        draw,
        rounded corners,
        fill=gray!15,
        font=\small,
        minimum width=6cm,
        minimum height=1.2cm,
        align=center
    },
    arrow/.style={
        ->,
        thick,
        >=stealth
    }
]
% Nodes
\node (src) {Source Code \\ Keystone, Kernel, Applications};
\node (buildroot) [below=of src] {Buildroot \\ Rootfs + Kernel};
\node (make) [below=of buildroot] {Make Orchestration \\ Keystone Runtime + Security Monitor};
\node (launch) [below=of make] {QEMU Launch};
\node (bench) [below=of launch] {Run Enclave / Benchmark};

% Arrows
\foreach \from/\to in {src/buildroot, buildroot/make, make/launch, launch/bench} {
    \draw[arrow] (\from) -- (\to);
}
\end{tikzpicture}
\caption{Build and deployment workflow from source code to running enclave and benchmark execution.}
\label{fig:build_deployment_workflow}
\end{figure}

@@

\begin{table}[h]
\centering
\begin{tabular}{l l p{6cm}}
\toprule
\textbf{Parameter} & \textbf{Default} & \textbf{Description} \\
\midrule
QEMU\_PORT & 9821 & SSH port forwarding to guest \\
QEMU\_DBG\_PORT & QEMU\_PORT+1 & TCP port for GDB debugging \\
QEMU\_RAM\_SIZE & 128M & Emulated RAM size \\
QEMU\_CORE\_COUNT & 2 & Number of virtual CPU cores \\
\bottomrule
\end{tabular}
\caption{Key QEMU launch parameters and defaults.}
\end{table}

@@

\begin{lstlisting}[language=bash,caption={Configuration options for cache simulation},label={lst:cache_config}]
# Instruction cache parameters
icachesize=SIZE         # Size of instruction cache
iblksize=BLOCK_SIZE     # Block size
iassoc=ASSOCIATIVITY    # Associativity

# Data cache parameters
dcachesize=SIZE         # Size of data cache
dblksize=BLOCK_SIZE     # Block size
dassoc=ASSOCIATIVITY    # Associativity

# Cache eviction policy
evict=lru|rand|fifo     # Replacement policy

# Logging configuration
limit=TOP_N             # Number of top thrashing instructions to record

# Monitoring settings
cores=N_CORES           # Number of cores to monitor
\end{lstlisting}

@@

\begin{lstlisting}[language=bash,caption={Kyber cryptographic API functions},label={lst:kyber_api}]
indcpa_keypair   # Key pair generation for IND-CPA variant
indcpa_enc       # Encryption (encapsulation) under IND-CPA assumptions
indcpa_dec       # Decryption (decapsulation) under IND-CPA assumptions

kyber_keypair    # Enhanced key pair generation supporting IND-CCA
kyber_encaps     # IND-CCA-compliant encapsulation
kyber_decaps     # IND-CCA-compliant decapsulation
\end{lstlisting}
Each of these operations was executed across the five representative system configurations defined in Section~\ref{sec:param-variation}, which span a broad range of memory capacities, core counts, and execution modes. To ensure measurement accuracy and reduce variability from transient system states, ten iterations were run per configuration with the first iteration discarded as a warm-up phase. Execution timing was obtained using the enclave’s internal high-resolution timer API, capturing secure-world runtime exclusively. In parallel, external system-level metrics — such as CPU usage, memory footprint, and context switching counts — were gathered from the host environment to provide contextual insights into system overhead and resource utilization.

@@

\begin{lstlisting}[language=bash,caption={Cache configuration profiles},label={lst:cache_profiles}]
# Small caches
Size: 4-8 KB
Associativity: Low
Eviction Policy: LRU

# Mid-sized caches
Size: 8-16 KB
Associativity: Mixed
Eviction Policy: FIFO or Random

# Large caches
Size: 32 KB
Associativity: High
Eviction Policy: FIFO
\end{lstlisting}

@@

The primary performance metric collected was total wall-clock execution time, measured from within the guest Linux environment. For the Kyber benchmark, high-resolution timing was implemented explicitly using the \texttt{clock\_gettime()} function, which was integrated into the benchmark source code prior to compilation. In contrast, both Dhrystone and CoreMark include built-in timing and throughput reporting mechanisms as part of their standard distributions. Accordingly, benchmark-specific performance metrics were used where applicable: Dhrystone MIPS (DMIPS) for Dhrystone, and iterations per second for CoreMark. These metrics enabled more nuanced comparisons of workload efficiency across various enclave and system configurations.

In addition to these high-level metrics, detailed system-level runtime statistics were gathered using standard Linux facilities such as the \texttt{getrusage()} system call. These included:

\begin{lstlisting}[language=bash]
# CPU time: Total time spent in user mode and kernel mode.
# Peak RSS: Maximum resident memory usage during execution.
# Page faults: Number of soft (minor) and hard (major) page faults.
# Context switches: Voluntary and involuntary context switching events.
\end{lstlisting}
